---
title: "Prull_Master_Analysis"
author: "Marcus Prull"
date: "12/10/2024"
output: html_document
---


Loading in required packages for analysis
```{r setup, include=FALSE}

require(lubridate)#For making date column
require(cowplot)#For any plot griding 
require(dplyr)#Filtering data
require(ggplot2)#Making final plots
require(geosphere)#package for straight line distance calculations
require(sf)#packing to read my spatial objects, EX. My Neely Henry Shapefile
require(raster)#used for raster functions
require(terra)#spatial analyst functions
require(gdistance)#transition function package
```

First I will start with analysis of our control individuals. Below we will calculate movement rates for both species as well as home range sizes. Additionally, we will look how movement rates differ throughout the year for each species. As well as compare straight line and in lake distance calculations (many studies in past have used straight line distance to represent movement)

Loading in raw control fish location data from github
```{r}

Control = read.csv("https://github.com/PrullMarcus/MastersAnalysis/raw/main/Data/ControlMovementFinal.csv")#had to change /blob/ to /raw/
Control

```

Here I filter the movement spreadsheet to fish that have >2 locations. The first row of data for each fish is its tagging location, therefore, we need 2 subsequent locations AFTER their tagging location to calculate a meaningful movement value. 
```{r}

Control = Control %>% group_by(RewardTag1) %>% filter (n()>2)#grouping chunks of data by their reward tag number and then filtering to only those groups that have 3+ observations

```

Here I make a singular date column for the dataset by using the year, month, and day columns from the raw csv
```{r}

Date = make_date(year = Control$Year, month = Control$Month, day = Control$Day)
Control = cbind(Date,Control)#binding the vector to the main csv
colnames(Control)[1]<-"Date"#Naming the date column "Date"
Control

```
Here I setup vectors to be used by the loop to ensure that it loops over each tag number independently of one another. Additionally, I setup the empty containers for the results of the loops to go into. One container is the raw distance values between successive locations and the second container is the days elapsed between successive observed locations. 
```{r}

tag_nums = unique(Control$RewardTag1)
n_fish = length(tag_nums)#90 individual fish making up our control dataset

control_SL = numeric()
control_SL_days = numeric()

```

First we will setup the loop to calculate STRAIGHT LINE DISTANCE between successive locations for my control individuals. 

```{r}
#Outside loop that selects tag-number 
for(i in 1:n_fish){
  
  new_tag = tag_nums[i]
  new_dat = subset(Control,RewardTag1 == new_tag)
  tmax = nrow(new_dat)
  ControlDist = numeric(tmax)
  days = numeric(tmax)
  
  for(t in 2:tmax){
    
    dist_m = geosphere::distm(c(new_dat$Long[t],new_dat$Lat[t]),c(new_dat$Long[t-1],new_dat$Lat[t-1]))#
    ControlDist[t] = dist_m
    days[t] = as.numeric(new_dat$Date[t]-new_dat$Date[t-1])
  }
  control_SL = append(control_SL,ControlDist)
  control_SL_days = append(control_SL_days,days)
}


```

Here we just combine the output from the loop to the original dataset. 
```{r}
Control = cbind(control_SL, Control)
colnames(Control)[1]<-"SL_M"#straight line_meters

Control = cbind(control_SL_days,Control)
colnames(Control)[1]<-"DaysElapsed"

Control
```
Now, because the first location is the original tagging location, the distance from the first to second record is essentially negligible. We didn't start tracking till almost 3 ish months after tagging. Therefore, the days elapsed and distance values from the first two rows for EACH fish is essentially negligible. Here I filter the data down to reflect that. 

Additionally, I add the movement rates (meters per day) as a new column to the data

```{r}

Control$SL_M_D = Control$SL_M/Control$DaysElapsed

```


Now we can calculate the in lake distances for these fish which takes a substantial time to run on the computer. The only thing that will change is the inputs into the loop. To do this I load in the Neely Henry Shapefile that I created on ArcGisPro from Github. I then convert the shapefile into a raster which is then used to create a "cost" raster. The cost raster is indicative of the water and excludes land. It then calculates the most conservative distance between successive points while staying within the bounds of the reservoir. 

The first thing we need to do is load in the shapefile

```{r}
Neely = st_read("/vsicurl/https://github.com/PrullMarcus/MastersAnalysis/raw/main/Data/NeelySHP_Jan2024)_NAD_1983_2011.shp") 
plot(Neely)
```
Now we need to convert the shapefile into a vector, then a raster, and then a transition matrix which is then used by the least cost distance functions. 

```{r}
NeelyVec = vect(Neely)
NeelyR = rast(NeelyVec, res = 0.00008)#res of 8.8 m grid cells, value of 1 is 1 "degree". that is why we need to use a big decimal to get it to meters. tried 00004 but it took way too long. 

NeelyRast = rasterize(NeelyVec,NeelyR)#value of 1 is Neely, NA value is land


NeelyRast[is.na(NeelyRast)] = 100


plot(NeelyRast)

```
Looks correct. Now I convert the cost raster into a transition layer (the input for the least-cost distance calculations) and then correct for corners. I have plotted the cost raster at the end.

```{r}

NeelyRast = raster(NeelyRast)#Converts spatraster to "raster" which is a different form of raster that the transition function needs in order to work properly. 
NeelyTrans = transition(1/NeelyRast, transitionFunction = mean, directions = 8)

#Here we correct for corners
NeelyTrans = geoCorrection(NeelyTrans, type = 'c', multpl=F)

NeelyTransRast = raster(NeelyTrans)
plot(NeelyTransRast)#looks correct

```

Before we run the loop we need to create a spatial object (points) from the lat/long columns from our control fish dataframe. The least cost function requires a spatial object as input into the calculation that has the same coordinate reference system as the Neely Shapefile. 

```{r}

Control = sf::st_as_sf(Control, coords = c("Long","Lat"),crs = crs(Neely))#Change to control(adds geometry but wont get rid of anything).

```

Here I check to make sure the points show up on the same map.
```{r}

plot(NeelyTransRast)
plot(Control,add=T)

```
They show up on the lake which is correct. If the CRS weren't matched up they wouldn't show up correctly. 

Here we do the same thing as before. The only thing that changes is the outputs are saved under a different name(LC = least cost) whereas before it was SL (straight-line). Additionally, the "distm" function is now changed to costDistance from the gdistance package in order to do least-cost distances (within lake distances). 
```{r}

tag_numbs = unique(Control$RewardTag1)#unique tag numbers
n_fish = length(tag_nums)#number of fish

control_LC = numeric()
control_LC_days = numeric()

for(i in 1:n_fish){
  
  new_tag = tag_nums[i]
  new_dat = subset(Control,RewardTag1 == new_tag)
  tmax = nrow(new_dat)
  ControlDist = numeric(tmax)
  days = numeric(tmax)
  
  for(t in 2:tmax){
    
    dist_m = gdistance:::costDistance(NeelyTrans, fromCoords = sf::st_coordinates(new_dat[(t-1),]),
                                                  toCoords = sf::st_coordinates(new_dat[t,]))
    ControlDist[t] = dist_m
    days[t] = as.numeric(new_dat$Date[t]-new_dat$Date[t-1])
  }
  control_LC = append(control_LC,ControlDist)
  control_LC_days = append(control_LC_days,days)
}
control_LC

```

I think the INF values have to deal with points that don't fall in a pixel or are not connected to the LC matrix. I am gonna remake the matrix and give land a super high cost so it won't go through land unless ABSOLUTELY NECESSARY.

```{r}

Control = cbind(control_LC, Control)
colnames(Control)[1]<-"LC_M"#least-cost meters
Control$LC_M_D = Control$LC_M/Control$DaysElapsed

```
#ControlFinal = Control %>% group_by(RewardTag1) %>% slice(3:n())

#ControlFinal$SL_M_D = ControlFinal$SL_M/ControlFinal$DaysElapsed

#ControlFinal

Now we need to make sure that we have data for only alive fish. The GroupMovement column is indicative of this. We need to make sure that the group movement values are all "A" so dead fish movement is not included. Every fish was considered to be "D" or "dead" for each subsequent location after its first "D" code. 
```{r}

ControlFinal = Control %>% group_by(RewardTag1) %>% slice(3:n())
ControlFinal = ControlFinal[ControlFinal$GroupMovement=="A",]

```

In total, we are left with 686 step lengths for fish that were not displaced by tournaments and were used for natural movement analysis

Firstly, lets look at how the straight-line movement rates look relative to the least cost rates.  

```{r}

#1:1 indicates complete agreement, this graph includes a few giant movements/outliers
plot(LC_M_D~SL_M_D, data = ControlFinal,ylim = c(0,2100), xlim = c(0,2100))
abline(a = 0, b = 1, col = "red")

#Zoomed in plot that is indicative of most movements <500 m/d
plot(LC_M_D~SL_M_D, data=ControlFinal, ylim = c(0,500), xlim = c(0,500))
abline(a = 0, b = 1, col = "red")

```
The above plot clearly shows the potential for underestimation of movement rates even at small movement values. 


Here is a boxplot for looking at how the monthly movement rates differed between species. 
```{r}

ControlFinal$MonthYear = strftime(ControlFinal$Date, format = "%Y-%m")
levels(ControlFinal$MonthYear)

#Boxplot comparing species movement over months. 
ggplot(ControlFinal,aes(as.factor(MonthYear),SL_M_D))+
      geom_boxplot(aes(fill=Species))+
      coord_cartesian(ylim=c(0,1600))+
      theme_classic(base_size=12)+
      scale_y_continuous(expand=c(0.013,0))+
      labs(x="Month-Year",y="Movement Speed (meters/week)")

```
Obviously this is hard to look at as there are many outliers across most of the months. Here I shrink down the plot and remove outliers to get a better look at what is going on with our movement across months.

Trying something to get the year and month to not overlap each other
```{r}
ggplot(ControlFinal,aes(as.factor(MonthYear),SL_M_D))+
      geom_boxplot(aes(fill=Species),outlier.color=NA)+
      coord_cartesian(ylim=c(0,250))+
      theme_classic(base_size=12)+
      scale_y_continuous(expand=c(0.013,0))+
      labs(x="Month-Year",y="Movement Speed (meters/week)")

```


######################################################################################
######################################################################################
######################################################################################

DISPERSAL STUFF 


First of all I will load in my dispersal data and then sort it to where the group movement variable is set to "A" to make sure we are only calculating movement for fish that are alive (same as we did for the control portion). I then created a spatial points class from the lat longs and then plotted them on the Neely shp file to make sure they show up on the map.
```{r}

Dispersal = read.csv("https://github.com/PrullMarcus/MastersAnalysis/raw/main/Data/DispersalFishFinal.csv")
Dispersal = subset(Dispersal, GroupMovement == "A")#subsetting to only alive fish
Dispersal = sf::st_as_sf(Dispersal, coords = c("Long","Lat"),crs = crs(Neely))#making spatial object out of lat/long columns

plot(Neely$geometry)
plot(Dispersal, add = T)

```
For tournament displaced fish their starting location is the boat ramp on the day they were caught/released and then every successive location was collected via manual tracking. Thus, we need to make sure that our dispersal fish have more than 1 observation to get a step lenght on them. Here I filter fish down to n>1. This removes fish that were reported as released in a tournament but we were never able to find afterwards. 
```{r}

Dispersal = Dispersal %>% group_by(RewardTag1) %>% filter(n()>1)

```


Here I make a date column which will be used to calculate days elapsed between step lengths as well as days elapsed since the fish was released at a tournament boat launch. 
```{r}

Dates = make_date(year=Dispersal$Year,month=Dispersal$Month,day=Dispersal$Day)
Dispersal = cbind(Dates,Dispersal)
colnames(Dispersal)[1] <- "Date"

```



Here is the loop that calculates each given fishes proximity to the ramp over time. Additionally, it calculates the movement rates/speeds of these fish over time. Both are stored in separate vectors. The proximity to the ramp is calculated by just doing a least cost distance of each location to each fish's starting location (the boat ramp). The movement rates/speeds are calculated in the same manner as they were for the control (non translocated fish) by calculating the step length between successive locations and then dividing by the time elapsed between the steps.Additionally, we add a vector that is WEEKS ELAPSED post capture to help with the first goal. 

```{r}


tag_numbs = unique(Dispersal$RewardTag1)#unique tag numbers
n_fish = length(tag_numbs)#number of fish

Disp_LC_Prox = numeric()
Disp_LC_Step = numeric()
Disp_Days_Step = numeric()
Disp_Days_Rel = numeric()

#control_LC = numeric()
#control_LC_days = numeric()

for(i in 1:n_fish){
  
  new_tag = tag_numbs[i]
  new_dat = subset(Dispersal,RewardTag1 == new_tag)
  tmax = nrow(new_dat)
  Dispersal_prox = numeric(tmax)
  Dispersal_step = numeric(tmax)
  Days_Step = numeric(tmax)
  Days_Rel = numeric(tmax)
  
  for(t in 2:tmax){
    
    prox_m = gdistance:::costDistance(NeelyTrans, fromCoords = sf::st_coordinates(new_dat[1,]),
                                                  toCoords = sf::st_coordinates(new_dat[t,]))
    step_m = gdistance::costDistance(NeelyTrans, fromCoords = sf::st_coordinates(new_dat[(t-1),]),
                                                  toCoords = sf::st_coordinates(new_dat[t,]))
    Dispersal_prox[t] = prox_m
    Dispersal_step[t] = step_m
    
    Days_Step[t] = as.numeric(new_dat$Date[t]-new_dat$Date[t-1])
    Days_Rel[t] = as.numeric(new_dat$Date[t]-new_dat$Date[1])
    
  }
  Disp_LC_Prox = append(Disp_LC_Prox,Dispersal_prox)
  Disp_LC_Step = append(Disp_LC_Step,Dispersal_step)
  
  Disp_Days_Step = append(Disp_Days_Step,Days_Step)
  Disp_Days_Rel = append(Disp_Days_Rel,Days_Rel)
}
Disp_LC_Prox
Disp_LC_Step
Disp_Days_Rel
Disp_Days_Step

```
```{r}

Dispersal = cbind(Disp_LC_Prox, Dispersal)
colnames(Dispersal)[1]<-"LC_Prox_m"#least-cost meters proximity to release ramp

Dispersal = cbind(Disp_LC_Step, Dispersal)
colnames(Dispersal)[1]<-"LC_Step_m"#least cost step length in meters

Dispersal = cbind(Disp_Days_Rel, Dispersal)
colnames(Dispersal)[1]<-"Days_Since_Rel"#Days since fish was initial released

Dispersal = cbind(Disp_Days_Step, Dispersal)
colnames(Dispersal)[1]<-"Days_Step"#Days between previous location and current location for which the step length was calculated

Dispersal$LC_M_D = Dispersal$LC_Step_m/Dispersal$Days_Step

Control$LC_M_D = Control$LC_M/Control$DaysElapsed

```

###### MAKE GRAPHS HERE 
###### PROX TO RAMP OVER TIME
###### MOVEMENT SPEEDS OVER TIME 
###### PERCENT WITHIN 3 km over time
###### ETC. 










######################################################################################
######################################################################################
######################################################################################

Here will will calculate homeranges for the fish that we have enough datapoints for. 


HOMERANGE STUFF 
















######################################################################################
######################################################################################
######################################################################################



TAGGING SIMULATION 